--- 
title: "Traits data in R"
author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [references.bib]
link-citations: yes
github-repo: TGuillerme/Morphological_traits
description: "A guide to using morphological trait data in R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble

## Requirements

### General computer level

In this workshop I will assume you are already familiar with how your computer works basically. You know how to do the following:

 * Open a file with a specific software (sometimes that's not just double clicking)
 * Locate a file in your computer (you know what a path is)

#### Required software

 * `R` ([the latest version](https://cran.r-project.org/))[^1]
 * A plain text editor (NOT WORD!)
 * *recommended:* an IDE ([integrated development environment](https://en.wikipedia.org/wiki/Integrated_development_environment)). For example [Rstudio](https://posit.co/products/open-source/rstudio/) or [Sublime Text](https://www.sublimetext.com/).

[^1]: **Always update your software**. If you do that already, thank you very much. If you don't, have a thought for the software developers that try very hard to correct/improve/update your favourite software. They don't do it for no reason (and no, it's not that hard to click on "update" button). 

### `R` level

I will assume you are already familiar with basic `R`. The basic notions that I'll assume you know are:

 * What is a package (e.g. `ape` or `dispRity`)
 * What is an object (e.g. `this_object <- 1`)
 * What is an object's class (e.g. the class `"matrix"` or `"phylo"`)
 * What is a function (e.g. the function `mean(c(1,2))`)
 * How to access function manuals (e.g. `?mean`)

#### Required packages

For this tutorial we will be using the packages `ape` (@ape), `dispRity` (@dispRity) and `treats` (@treats):

```{r, eval = FALSE}
required_packages <- c("ape", # for analysing phylogenetic data
                       "dispRity", # for disparity analyses
                       "treats") # for some trees and traits plots
install.packages(required_packages)
```

Note that you can also use the excellent following packages for more specific tasks not covered here:

 * `Claddis` for analysing discrete morphological characters (@lloyd2018journeys)
 * `paleotree` for analysing palaeo phylogenetic data (@bapst2012paleotree)
 * `strap` (@strap) or `deeptime` (@deeptime) for plotting nice geological time scales
 * `palaeoverse` (@palaeoverse) for wrangling palaeo data
 * `Morpho` (@morpho) or `geomorph` (@geomorph) for analysing geometric morphometric data

Let's get into it.

## Glossary

Trait data is going to be the generic term here to designate any type of data that describe some measurable features of a specimen, species, group, etc.
Here is a glossary table from @moms that specify what I mean throughout this book (and check @mammola2021concepts if you're more coming from ecology).

Mathematics | Ecology | Macroevolution | This book
--------------------|--------------------|--------------------|------------
Matrix ($n \times d$) with a structural relation between rows and columns | Functional space, morphospace, etc. | Morphospace, traitspace, etc. | traitspace
Rows (*n*) | Taxa, field sites, environments, etc. | Taxa, specimen, populations, etc. | observations
Columns (*d*) | Traits, Ordination scores, distances, etc. | Traits, ordination scores, distances, etc. | dimensions
Matrix subset ($m \times d$; $m \leq n$) | Treatments, phylogenetic group (clade), etc. | Clades, geological stratum, etc. | group
Statistic (i.e. a measure)  | Dissimilarity index or metric, hypervolume, functional diversity, etc.  | Disparity metric or index | space occupancy measure
Multidimensional analysis | Dissimilarity analysis, trait analysis, etc. | Disparity analysis, disparity-through-time, etc. | multidimensional analysis

Note that here traits are treated as unidimensional (i.e. one column = one trait) but it doesn't need to be the case!
For example you can have a multidimensional trait that is described using multiple numbers (a multidimensional trait), for the trait "location" can be described using both latitude and longitude (a 2D trait).

## How does this work?

Throughout the manuscript you will see some different coloured zoned that you can use to either to load your own data (in blue), catch up if you feel a bit behind (orange), or play around with different options (green).

### Use your own data

<div class="warning" style='padding:0.1em; background-color:#5D7BBB; color:#092663'>

**USE YOUR DATA**: load your own data!

```{r, eval = FALSE}
## Read your favourite csv file (you'll need to set up the path yourself!)
continuous_data <- read.csv("path_to_my_super_serious_dataset.csv")
```
</div>


### Catching up

<div class="warning" style='padding:0.1em; background-color:#FFAF70; color:#954200'>

**CATCHING UP ZONE**: loading some example data!

```{r}
## Loading some data
data(mtcars)
## Giving the data the correct name so that you can follow
my_data <- mtcars
```
</div>

### Tinker time!

<div class="warning" style='padding:0.1em; background-color:#65D85F; color:#067800'>

**TINKER TIMES**: play around with this very complicated plot.

```{r}
## Plot data
plot(my_data[, c(1,3)], col = "black", pch = 19, main = "booooring")
```

Can you find a solution to make it less boring?

<details>
  <summary>[**Click to expand the solution**]:
</summary>

```{r}
## Plot data
plot(my_data[, c(1,3)], col = rainbow(nrow(my_data)), pch = 19, main = "better?")
```

</details>
</div>



## Help me!

Whether you are doing this workshop online or in person, please help me improve this whole workshop:

 * Something is not working
 * Something is unclear
 * There's a typo (realy)!

You can just shout at me or [drop me an email](mailto:t.guillerme@sheffield.ac.uk)!

## References